---
title: "pca fresh start"
author: "Wayne Monical"
date: "4/28/2021"
output: html_document
---


Notes:
all.hap has a good linear scree plot ( I was wrong. That wall all of them)
also a goal: conduct PCA on the original and the square-rooted data (DONE, with similar results as others)
What I need to do is rescale the drug treatment levels





Library Load
```{r}
library(readr)
library(tidyverse)
library(scatterplot3d)
library(kernlab)
```



```{r}
# info load
experiment.information <- read.csv("experiment.information.csv")

# BIG CHANGE
# We're losing regression strength from the fact that mock and unt are their own catergories

experiment.information$trtmnt.type[experiment.information$trtmnt.type == "unt"] <- "Gy"
experiment.information$trtmnt.type[experiment.information$trtmnt.type == "mock"] <- "Ola"


all.reads <- read.csv("all.crispresso.clean.csv")
all.reads <- all.reads[,-1]

replace.list <- as.list(rep(0, ncol(all.reads)))
names(replace.list) <- names(all.reads)


# replace NA's with zeros
all.reads <- replace_na(data = all.reads, replace = replace.list)

```




```{r}
# WT will only be in all.reads
# p4 and p5 will only be in the all.reads, all.hap, and all.dip
cell.type <- experiment.information$Cell.Type
treatment.type <- experiment.information$trtmnt.type

# percent.boolean takes care of the NA's                                                                     # drops WT's             # drops p4's and p5's
percent.boolean <- experiment.information$data.type == "%" & !is.na(experiment.information$data.type)        & cell.type != "WT"      & treatment.type != "p4" & treatment.type != "p5"




all.boolean <- percent.boolean
all.hap <- cell.type == "H" & percent.boolean
all.dip <- cell.type == "D" & percent.boolean

hap.nira <- cell.type == "H" & percent.boolean & treatment.type == "Nira"
hap.ola <- cell.type == "H" & percent.boolean & treatment.type == "Ola"
hap.gy <- cell.type == "H" & percent.boolean & treatment.type == "Gy"

dip.nira <- cell.type == "D" & percent.boolean & treatment.type == "Nira"
dip.ola <- cell.type == "D" & percent.boolean & treatment.type == "Ola"
dip.gy <- cell.type == "D" & percent.boolean & treatment.type == "Gy"

boolean.list <- list(all.boolean = all.boolean, all.hap = all.hap, all.dip = all.dip, hap.nira = hap.nira, hap.ola = hap.ola, hap.gy = hap.gy, dip.nira = dip.nira, dip.ola = dip.ola, dip.gy = dip.gy)
```



To do for the 10000th time: turn NA's to zeros in all.reads
```{r}
mutations.na <- apply(all.reads[,all.dip], 1, function(x){sum(is.na(x))}) > 0
summary(mutations.na)

view.reads <- all.reads[order(all.reads$`By mutation(s)`),]

length(unique(all.reads$`By mutation(s)`))
length(all.reads$`By mutation(s)`)

```


SUCESS!!!
```{r}
# boolean.vector is logical as to whether the ith of n data points should be included in the analysis

begin.object <- function(boolean.vector, y.data, x.set){
  list(y.data = y.data[,boolean.vector], x.set = x.set[boolean.vector,])
}

conduct.linear.pca <- function(list.object){
  
  # checks
  if(prod(names(list.object) != c("y.data", "x.set"))){
    stop("incorrect list.object. Use begin.object function")
  }
  
  
  linear.pca <- prcomp(t(list.object$y.data), center = FALSE)
  
  c(list.object, linear.pca = list(linear.pca))
  
  
}



# testing
first.obj <- begin.object(boolean.vector = all.boolean, y.data = all.reads, x.set = experiment.information)
first.obj.pca <- conduct.linear.pca(list.object = first.obj)

plot(first.obj.pca$linear.pca)
```



Applying it to all booleans
```{r}
all.obj <- lapply(boolean.list, FUN = function(x){begin.object(boolean.vector = x, y.data = all.reads, x.set = experiment.information)})


# it literally works
all.obj.lpca <- lapply(all.obj, FUN = function(x){conduct.linear.pca(x)})


# lets try to use the screeplot function
# none of these screeplots are good except the one for all the data
lapply(all.obj.lpca, FUN = function(x){screeplot(x$linear.pca)})
```

Plots to save
```{r}
par(mfrow = c(1,2))
screeplot(all.obj.lpca$all.hap$linear.pca, main = "Screeplot of All Haploid Experiments")
screeplot(all.obj.lpca$all.dip$linear.pca, main = "And of All Diploid Experiments")

```



Conduct Regression

I'd like conduct regression to take either linear of kernel pca's
Well, I'd like it to take the whole object
```{r}
conduct.regression <- function(list.object, linear.pca = TRUE, anova = FALSE){
  
  # lm needs a formula
  regression.formula <- (formula.function(list.object))[[1]]
  
  if(linear.pca){
    
    # I'm putting this chunk on hold
    # Has linear pca been conducted? If not, conduct it
    #if(! linear.pca %in% names(list.object)){
     # list.object <- conduct.linear.pca(list.object = list.object)
    #}
    
    y.set <- list.object$linear.pca$x[,1]
    
    
    
    design.matrix <- data.frame(y.set = y.set, list.object$x.set)
    
    lpca.regression <- lm(formula = regression.formula, data = design.matrix)
    
    if(!anova){
      return(c(list.object, lpca.regression = list(lpca.regression)))
    }
  }
  
  
}


conduct.regression2 <- function(list.object, pcas = 1:3){
  
  # lm needs a formula
  regression.formula <- (formula.function(list.object))[[1]]
  
    
  num.list <- create.numlist(pcas)  
    
  lpca.regressions <- lapply(num.list, FUN = function(y){apply.regression(y.set = list.object$linear.pca$x[,y],
                                                                          x.set = list.object$x.set,
                                                                          formula = regression.formula
                                                                          )})
  return(c(list.object, lpca.regressions = list(lpca.regressions)))
  
}



apply.regression <- function(y.set, x.set, formula){
  design.matrix <- data.frame(y.set = y.set, x.set)
  lm(formula = formula, data = design.matrix)
  }



create.numlist <- function(integer.vector){
  names(integer.vector) <-  integer.vector
  as.list(integer.vector)
}



# conduct.regression depends on formula.function
formula.function <- function(list.object){
  cell.type.include <- length(unique(list.object$x.set$Cell.Type)) > 1
  treatment.type.include <- length(unique(list.object$x.set$trtmnt.type)) > 1
  
  possible.formulas[cell.type.include  + treatment.type.include + 1]
}



# formula.function depends on this constant
possible.formulas <- c(y.set ~ trtmnt.amnt, y.set ~ trtmnt.type + trtmnt.amnt, y.set ~ Cell.Type + trtmnt.type + trtmnt.amnt)


# test. It works!
first.obj.reg2 <- conduct.regression2(first.obj.pca)
summary(first.obj.reg2$lpca.regressions$`1`)



```

Doing all regressions

```{r}
all.obj.reg2 <- lapply(all.obj.lpca, FUN = function(z){conduct.regression2(z, pcas = 1:4)})


lapply(all.obj.reg2$all.hap$lpca.regressions, FUN = summary)
```



Looking at all regressions
```{r}
all.regressions <- lapply(all.obj.reg2, FUN = function(w){lapply(w$lpca.regressions, FUN = summary)})
all.regressions

```

```{r}
y.set <- all.obj.reg2$all.hap$linear.pca$x[,1]
x.set <- all.obj.reg2$all.hap$x.set
ex.mat <- data.frame(y.set = y.set, x.set)
ex.lm <- lm(y.set ~ trtmnt.amnt, data = ex.mat)
summary(ex.lm)
```



```{r} 
# This is the haploid diploid split
plot(all.obj.reg2$linear.pca$x[,1])

# this is how genes are effected
# 

plot(all.obj.reg2$all.hap$linear.pca$rotation[1:50, 1])


el.2nd.val <- all.obj.reg2$all.hap$linear.pca$rotation[1:50,2] * all.obj.reg2$all.hap$lpca.regressions$`2`$coefficients[4]
text.index <- rep(NULL, 50)
text.index[abs(el.2nd.val) > 0.1 ] <- all.reads$`By mutation(s)`[1:50] [abs(el.2nd.val) > 0.1]

plot(1:50, el.2nd.val, main = "First 50 Elements of The Second Principle Component", xlab = "Index", ylab = "Percent Change Coresp. to 1-Unit Increase in Dose")  + text(1:50, el.2nd.val, text.index, cex = 0.8, pos = 4)

# comparted to a regression
plot(1:50, el.2nd.val,  xlab = "Index", ylab = "Percent Change Coresp. to 1-Unit Increase in Dose", main = "PCA Slopes Versus Individually Fit Slopes")  +points(1:50, unlist(first.50.regressions), col = "red")




plot(1:50, el.2nd.val, main = "Slope Term of Treatment Amount Assigned to Each Mutation", xlab = "Index", ylab = "Percent Change Coresp. to 1-Unit Increase in Dose")  + text(1:50, el.2nd.val, text.index, cex = 0.8, pos = 4)


















plot(all.obj.reg2$all.hap$linear.pca$rotation[1:50,3])
plot(all.obj.reg2$all.hap$linear.pca$rotation[1:50,4])

hist(all.obj.lreg2$all.boolean$linear.pca$x[,1])



hist(all.obj.lreg2$all.hap$linear.pca$x[,1])
hist(all.obj.lreg2$all.hap$linear.pca$x[,2])
hist(all.obj.lreg2$all.hap$linear.pca$x[,3])


plot(all.obj.reg2$lpca.regression$residuals)
```







This might evaporate everything, but we have to try rescaling the treatment values, 2 to 5?



List index
y.set DONE
x.set DONE
linear.pca DONE
linear.pca.summary # summary you can just apply; that's not that big a deal
linear.pca.regression DONE
linear.pca.regression.summary


anova.pca
anova.pca.more
anova.pca.regression
anova.pca.regression.summary


I have right now
boolean.vector
y.data
x.set



nnet.plan
  starting point
    y.data
    x.set
    boolean
    
  primary process
    y.data[boolean]
    x.set[boolean]
    
  unstacking step
    y.unstacked
    x.set
    
  nnet step
    


Defending my choices
  Using all data points
    The facts: the screeplot looks better than it has any right to, since the variance you see is almost entirely the hap/dip split
    But the rest of the experiments only have 6 or so examples. We need the power of the tens-scale to exctract the info

  The multiple testing bonferroni correction works in our favor. 
    i.e. the bonferroni correction allows us to ignore non *** results
    124 params, our significance boundary becomes 
    
    
  Regressing on 2 Principle Components
    IT'S THE INTERCECPT AND THE SLOPE!!!!!!!!
  
  With the number of reads so high